---
title: '<div style=''text-align: justify''> How are COVID deaths correlated with Human
  Development Index? Whether high or low HDI has linear relationship with the fatality
  rate? </div>'
author: "Louis, Floriann Deepika"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 1
    number_sections: yes
    theme: united
  pdf_document:
    toc: yes
    toc_depth: '1'
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DataExplorer)
library(caret)
library(dplyr)
library(corrplot)
library(ggplot2)
library(ggpubr)
library(skimr) #SKIM function - For statistical summary
library(EnvStats) #Roseners test for outliers
library(gclus) #Scatterplot Matrices from the glus Package 
library(scatterplot3d)  #Visualization
library(plotly) #world map
library(readxl) #Read UN dataset
library(superml)
library(RColorBrewer) #Bar chart
library(base64)
suppressWarnings
suppressMessages
options(error = expression(NULL))
```

# Motivation
<div style="text-align: justify">
The COVID-19 pandemic has not only affected global economies in terms of
GDP growth, but also on the path of human development. The core
foundations of the concept of human development i.e. a healthy and long
life has also been affected. Human development is measured using three
core indicators which are; life expectancy (at birth,) knowledge (mean
years of schooling) and standard of living (Gross National Income per
capita.) The Human Development Index is a critique of the 'economic
growth before people' model of development. In this project I would
like to answer the question: How COVID-19 related deaths correlates with
the Human Development Index (considering both low and high HDI
countries?) This topic peaked our interests as the media has speculated
that most developed countries who rank very high on the HDI topped the
charts of corona virus deaths worldwide per million population, albeit
having a strongest and most efficient healthcare facilities. In
contrast, it is believed that developing countries like India and South
Africa with middle-low ranked HDI Indices have managed to keep COVID-19
infections and death rates at some of the lowest levels in the world. To
check the relevancy of these speculations, I will use the Our World In
Data Covid data set to analyse Covid cases against the components of HDI
and provide a summary of findings.
</div>

# Analysis Plan
<div style="text-align: justify">
To answer our research question, Firstly, I will perform data retrieval
(identifying data sources, defining important variables and relevant
formulas for calculating Human Development Index). Secondly, I will
perform data processing which includes data transformation, data
translation, checking consistency of data, data pre-processing, checking
outliers and handling of missing values. Thirdly, I will perform
Exploratory Data Analysis, make statistical summaries, visualization of
results and correlation plots. Then I will perform a hypothesis testing
of HDI vs total deaths in (all countries, both low & high countries,
continent wave 1) and make a hypothesis comparison table as well as data
normalization. To confirm the hypothesis test result I will perform
linear regression for high HDI countries and identify the significant
variables which contributes more to covid death.
</div>

# Data retrieval

## Data source Identification:
<div style="text-align: justify">
Two datasets are considered for this analysis, namely OWID (Our World in
Data) and UN (United Nations).

Our World In Data (OWID): OWID could be the right data source for this
analysis. It auto extracts data from (Johns Hopkins University)JHU and
World Health Organisation(WHO) which are in turn getting daily data from
respective country health authorities and they are reliable. It has all
the required variables for analysis starting from HDI and its attributes
and total death per million. It contains totally 131289 observations
with 65 variables.

*Link:
<https://github.com/owid/covid-19-data/raw/master/public/data/owid-covid-data.csv>*

United Nations (UN): UN data contains all the statistical information
specific to country. I have chosen UN data to extract the education
index which is not available in the OWID dataset. It contains totally
195 observations with 15 variables.

*Link: <http://hdr.undp.org/en/data>*
</div>

## Data retrieval/data extraction

OWID data set is extracted from the database and saved as data_Covid.
Similarly, UN data set is extracted and saved as un_edu_data

```{r, results='hide',message=FALSE}
#Extracting the OWID dataset:
if (!dir.exists("data")) { dir.create("data") } 
if (!file.exists("data/owid-covid-data.csv")) 
  download.file("https://covid.ourworldindata.org/data/owid-covid-data.csv",destfile = "data/owid-covid-data.csv")
data_Covid <- read_csv("data/owid-covid-data.csv", 
                       col_types = cols(
                         .default = col_double(),  
                         iso_code = col_character(),  
                         continent = col_character(),
                         location = col_character(),  
                         date = col_date(format = ""), 
                         tests_units = col_character()
                       )) 

#Extracting UN dataset for education index:
if (!file.exists("data/un_edudata.xlsx")) 
  download.file("http://hdr.undp.org/sites/default/files/2020_statistical_annex_all.xlsx",destfile ="data/un_edudata.xlsx",
                mode = "wb")
un_edu_data <- read_xlsx("data/un_edudata.xlsx", sheet = 2, range = "A5:O200")
```

## Define the variables
<div style="text-align: justify">
Feature selection is one the vital task to perform data analysis and
exploration There are various statistical methods included in feature
selection. In this project our motive is to focus on the Human
Development index and the total death. Hence, I select features which
are closely related to this component. The OWID dataset has 65
variables. All the variables are not required for our analysis. I
select only few variables here,

**gdp_per_capita:** Per capita income or total income measures the
average income in a given country in a specified year.It is calculated
by dividing the country's total income by its total population size.OWID
doesn't mention the source, but this data point is maintained by World
bank. Latest available is 2011 constant USD. The values are in fraction.
It ranges from 702 in Burundi/Africa to 116935 in Qatar/Asia. Units are
measured in USD. This variable takes the purchasing power parity into
account.

**Life expectancy:** Life expectancy is the average age an individual is
expected to live based on statistical average. Central African Republic
has low life expectancy of 53.28 and Monaco has highest range of 86.75.
The values are in fraction. The observations are extracted from WHO into
OWID.

**Education Index:** It refers to the knowledge received through
schooling or instruction and to the institution of teaching as a
whole.The values of the index are in fraction. Seychelles has the
highest mean years of schooling with HDI rank of 67 and Burkina Faso has
the lowest mean years of schooling with HDI rank of 182.

**human_development_index:** The Human Development Index is a statistic
composite index of life expectancy/long-healthy life,
education/knowledge, and per capita income/standard of living, which are
used to rank human development of the countries. This is calculated in 2
steps from 3 variables for each countries (life expectency / Health,
Education/Expected & Mean years of schooling, Standard of living/GDP Per
capita PPP). Values of each of the four metrics are first normalized to
an index value of 0 to aggregating the four metrics to produce the HDI.
HDI is 1 if a country achieves maximum value(well developed) and it is 0
for a country that is at the minimum value

*Reference Link 1 : OWID
<https://ourworldindata.org/human-development-index>*

*Reference Link 2 : United Nations
<http://hdr.undp.org/en/indicators/137506>*

**Total_death_per_million:** This defines the total confirmed deaths per
1, 000,000 people. The values are denoted in fraction. Here, Asia- India
has lowest total deaths per million as 0.001 and South America - Peru
having highest death rate of 66112.85 per million.
</div>

## Define HDI and its components

The Human Development Index (HDI) is a statistic developed and compiled
by the United Nations to measure various countries levels of social and
economic development. The index is used to follow changes in development
levels over time and to compare the development levels of different
countries.

The **HDI components** comprise of:

1) Education -- mean years of schooling and expected years of schooling
and measured by adult literacy rate

2) Life expectancy at birth -- as an index of population health and
longevity

3) Gross national income per capita (GDP) -- it is a monetary measure of the finished goods in a country at a specific period.

## HDI formula and calculation
<div style="text-align: justify">
HDI is a single index measure which takes into account the 3 components
- extended healthy life, exposure to knowledge and a good standard of
living. It uses 4 matrices namely,

1\. life expectancy at birth

2\. Education

-   Expected years of schooling

-   Average years of schooling

3\. Gross National Income (GNI) per capita

Calculation is done using two steps:

Step 1: Formulating indices for each of the metrics Taking actual value
for a country, global maximum values and global minimum value is
evaluated using the formula,

Dimension index = (actual value - minimum value) / (maximum value -
minimum value)

Dimension index ranges between 0 and 1, which denotes minimum and
maximum value respectively.

Note: While calculating for knowledge, dimension index for expected
years of schooling is calculated separately and dimension index for
average years of schooling is calculated separately. Finally, arithmetic
mean of both the values are taken.

Step 2: Summing up the values of four metrics to generate HDI Having the
3 dimension indeces, now geometric mean of all the 3 values are taken in
order to calculate HDI using the following formula,

HDI = (I(Health) \* I(Education) \* I(Income))\^ (1/3)

*Link:
<https://ourworldindata.org/human-development-index#how-is-the-human-development-index-calculated>*
</div>

# Data processing

## Data Transformation

### Data translation

```{r}
#Selecting only required variables in data set with date filtered as maximum date which takes the updated date.
data_owid <- data_Covid %>% 
  select(continent, location, date, population,  median_age, aged_65_older, aged_70_older, gdp_per_capita, 
        extreme_poverty,cardiovasc_death_rate, diabetes_prevalence, female_smokers, male_smokers,
        handwashing_facilities, hospital_beds_per_thousand, life_expectancy, human_development_index, 
        total_deaths_per_million,total_deaths,iso_code) %>% 
        group_by(location) %>% filter(date == max(date)) %>% # Take the most recent date per country
        group_by() #removing groupby
```

<div style="text-align: justify">
I select only the required predictor variables with latest date which
will help us to achieve our goal of find the linear dependency between
HDI and total COVID deaths. Since the raw data is already in csv format,
there is no data transformation performed here.
</div>

```{r}
#Checking Number of rows and columns in the dataset:
dim(data_owid)
```

OWID dataset has 237 observations with 20 variables. UN dataset has 195
observations with 15 variables.

```{r}
#Summary statistics
summary(data_owid)
```

<div style="text-align: justify">
Summary provides details about the statistical analysis of each of the
features. It provides Minimum value, 1st Quantile, Median, Mean, 3rd
Quantile, maximum values and the number of NA values to validate our
data set. With respect to our response variable, I can observe that
there is huge difference between the minimum and maximum value of the
variables as 3.101 and 6006.490 respectively. This shows how widely our
data is spread and how extreme our observations are.
</div>

```{r}
#Removing missing values in y variable - total_deaths
data_owid <- data_owid %>% select(continent: iso_code)%>% 
              filter(!is.na(total_deaths_per_million))
```

<div style="text-align: justify"> 
I select total_deaths_per_million as our target variable. While
checking for the NA values, it shows that there are 39 values which are
NA. It is advisable to remove the NA values completely rather then
performing analysis with the missing values in the dependent variable
which might result in biased output.
</div>

```{r}
#Clean UN education dataset and calculate index variable:
#Rename education variables
names(un_edu_data)[2] <- "location"
names(un_edu_data)[7] <- "expected_yr_Schooling"
names(un_edu_data)[9] <- "mean_yr_Schooling"

#Filter and select only the relevant education variables
un_edu_data <- un_edu_data %>%
            select(location,expected_yr_Schooling,mean_yr_Schooling) 
un_edu_data <- un_edu_data %>%
            filter(!is.na(location)) %>% 
            filter(location !="Country") %>% 
            filter(!is.na(expected_yr_Schooling))
```

```{r}
#Computing Education index from UN data set:

#Change col type as numeric for finding average
un_edu_data$expected_yr_Schooling <- as.numeric(as.character(un_edu_data$expected_yr_Schooling))  
un_edu_data$mean_yr_Schooling <- as.numeric(as.character(un_edu_data$mean_yr_Schooling))  
sapply(un_edu_data, typeof)

#Calculate Education index
un_edu_data <- un_edu_data %>%
  mutate(education_index = (expected_yr_Schooling + mean_yr_Schooling)/2) %>%
  select(location, education_index)
apply(is.na(un_edu_data),2, sum)
```
<div style="text-align: justify"> 
Education index is calculated by summing up the expected year schooling
and mean year schooling and diving it by 2. These are done in the UN
data set as I will be using this variable further in our project as it
is the component of HDI.
</div>

## Data Pre-processing

### Checking Outliers
<div style="text-align: justify"> 
Outlines are extreme values that lies outside the minimum and maximum
range. They can be always removed from the data set as they might affect
the distribution of the data. It might cause error variance and thereby
reduce the performance of the various statistical tests which will be
performed further. Additionally, they also change the assumption that
the variables are in linear relationship with each other. There are
various ways in which outliers can be identified and handled.
</div>
```{r fig.align="center",fig.width = 16,fig.height=8}
#Visualizing outliers with box plot
ggplot(data_owid) +
  aes(x = "", y = total_deaths_per_million) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
```

<div style="text-align: justify"> 
The box plot displays how the variable - total_deaths_per_million is
dispersed with five components,

1\. Minimum

2\. First quartile (Q1)

3\. Median

4\. Third quartile (Q3)

5\. Maximum

From the box plot, I can observe that there is an outlier toward the
range of 6000. The minimum observation and the maximum observation in
the total death per million lies between 3.101 to 6006.490, with median
towards the 1st quartile. This shows that the data points are right
skewed. This means a wider range of observations of total deaths per
million are more spread between the median and the 3rd quartile.
Moreover, the data are condensed towards the 1st quartile and the
median.

Statistical test for detecting outliers:

There are three formal techniques which is used to detect outliers.

1\. Grubbs's test: This test identifies if the lower observation or the
highest observation is an outlier in a dataset

2\. Dixon's test: This is similar to Grubb's test. Once the outlier is
identified, this test will be performed on the identified outlier.

3\. Rosner's test: This test can be performed to identify multiple
outlier at the same time, which Grubbs's test and Dixon's test fails to
perform. This test also has an advantage to avoiding masking, where an
outlier which lies close to another outlier can go unidentified.
</div>

```{r}
#Rosner’s test - total_deaths_per_million
data_owid_death_out <- rosnerTest(data_owid$total_deaths_per_million,
                                  k = 3)
data_owid_death_out
data_owid_death_out$all.stats
```
<div style="text-align: justify"> 
The result of Rosner's test states that there is one outlier identified
in lines 142 which has a value of 6006.490 with a mean of 901.2832 and
standard deviation of 970.2075.
</div>
```{r}
data_owid_total_deaths_mill <- data_owid[order(-data_owid$total_deaths_per_million),] 
head(data_owid_total_deaths_mill,2)
```

The above tables states that Peru has the highest number of death which
is 6006.490 and Tanzania has lowest death of 3.101

```{r}
#Removing outliers from total_deaths_per_million:
data_owid <- data_owid %>% filter(total_deaths_per_million < 4000)
```

```{r fig.align="center",fig.width = 16,fig.height=8}
#Visualize the log transform of the total death variable:
ggplot(data_owid) +
  aes(x = total_deaths_per_million) +
  geom_histogram(bins = 30L, fill = "#0c4c8a") +
  theme_minimal() +
  xlab("Total Number of Deaths per million")+
  ggtitle('Histogram of Logarithm of Total Number of Deaths per million')
```
<div style="text-align: justify"> 
The above histogram illustrates that the outlier is removed.

### Handling missing values

**Missing values:** Missing values are observation which has no data.
Processing it without taking care of it might lead to poor performance
of the statistical test and machine learning models. It might lead to
more biased variable leading to inaccurate results.

Missing values are of 3 types.

1\. Missing completely at random (MCAR) : There is no relationship
between the observed and unobserved values which are missing.

2\. Missing at random : There is a relationship between the observed and
unobserved values as there exist no pattern between them.

3\. Missing not at random : The missing data is related to the
unobserved data. 4. Structurally missing data: There is a logical reason
behind the non-existance of data.

Approaches to handle missing data:

1\. Deletion methods to remove missing values.

2\. Regression analysis to systematically eliminate data.

3\. Data imputation techniques.
</div>

```{r}
#Check NA values in the dataset:
apply(is.na(data_owid),2, sum)
#Checking the percentage of the missing values:
p <- function(x) {sum(is.na(x))/length(x)*100}
apply(data_owid, 2, p)
```

There is a general assumption that 25 to 30% of missing values are
allowed. it is advisable to discard it if the variable is insignificant.

```{r}
#Treating NA values in continent variable:
data_owid <- data_owid %>% select(continent: iso_code) %>% 
  group_by(location) %>% filter(date == max(date)) %>% # Take the most recent date per country
  group_by() #%>% # Remove the grouping
data_owid <- data_owid %>% 
  filter(!is.na(continent))
```
<div style="text-align: justify"> 
Comparing the NA values of continent with the location, it can be
observed that the location has insignificant values named as world, high
income, low income. This shows that missing values in the continent
variable are Structurally missing data. Hence these values can be
dropped for further analysis.
</div>

```{r}
#Treating NA values in HDI variable:
data_owid_hdi <- data_owid %>% select(location, human_development_index)%>% 
  filter(is.na(human_development_index))
data_owid <- data_owid %>% 
  filter(!is.na(human_development_index))
```
<div style="text-align: justify"> 
There are 5 countries which has missing values namely Kosovo, Monaco,
San Marino, Somalia and Taiwan. Since it was only 5 countries, measures
was made to fill the missing value from the UN dataset. However, it was
noticed that HDI value is missing for these 5 countries in the UN
dataset itself. Hence dropping the NA values and proceeding with the
further analysis.
</div>
```{r}
#Treating NA values for median_age, aged_65_older and aged_70_older variables:
#Imputing median_age, aged_65_older and aged_70_older with mean
mean_imputer <- function(x) 
{x = round(replace(x, is.na(x) , mean(x, na.rm = TRUE)), digits=0)}
data_owid <- data_owid %>% mutate_at(5:7, mean_imputer)

#Treating NA values for handwashing_facilities variable:
#Removing handwashing_facilities since it has more than 50% of the data missing
data_owid <- data_owid %>% 
  select(-handwashing_facilities)
head(data_owid,2)

#Treating NA values for female_smokers and male_smokers with median:
median_imputer <- function(x) 
{x = round(replace(x, is.na(x) , median(x, na.rm = TRUE)), digits=0)}
data_owid <- data_owid %>% mutate_at(12:13, median_imputer)
```

```{r}
#Treating NA values for extreme_poverty :
data_owid_extPov <- data_owid %>% 
  select(location, extreme_poverty) %>% 
  filter(is.na(extreme_poverty))
data_owid <- data_owid %>% 
  select(-extreme_poverty)
```
<div style="text-align: justify"> 
Extreme poverty is an vital element which contribute to the GDP of a
country which shows an adverse effect on HDI. The missing dimensions
were approached to be filled with the data available in the world bank
dataset. While analyzing the missing values in the data set, it was
observed that values are missing for both countries with high HDI and
low HDI. This scenario can be considered as Missing Missing completely
at random (MCAR). Missing values were approached to be filled from the
world bank dataset, but it was found that most of the countries does not
have extreme poverty value recorded in the world bank itself. Here I
decided to drop the variable because treating the missing values with
imputation method might cause more bias which is not a wise approach.
</div>

```{r}
#Treating NA values for hospital_beds_per_thousand by imputing median:
data_owid_bed <- data_owid %>% select(location, hospital_beds_per_thousand)%>% 
  filter(is.na(hospital_beds_per_thousand))
data_owid <- data_owid %>% mutate_at(13, median_imputer)

#Treating NA values for gdp_per_capita by imputing mean:
data_owid_gdp <- data_owid %>% select(location, gdp_per_capita,human_development_index) %>% 
  filter(is.na(gdp_per_capita))
data_owid <- data_owid %>% mutate_at(8, mean_imputer)

#Treating missing values for cardiovasc_death_rate anddiabetes_prevalence by imputing median:
data_owid_cardio <- data_owid %>% select(location, cardiovasc_death_rate) %>% 
  filter(is.na(cardiovasc_death_rate))
data_owid_diabetes <- data_owid %>% select(location, diabetes_prevalence) %>% 
  filter(is.na(diabetes_prevalence))
data_owid <- data_owid %>% mutate_at(9:10, mean_imputer)
```

### Joining dataset - UN and OWID

```{r}
#Replacing the country name of UN dataset with OWID for joining both the datasets 
un_edu_data$location[un_edu_data$location == "Viet Nam"] <- "Vietnam"
un_edu_data$location[un_edu_data$location == "Venezuela (Bolivarian Republic of)"] <- "Venezuela"
un_edu_data$location[un_edu_data$location == "Timor-Leste"] <- "Timor"
un_edu_data$location[un_edu_data$location == "Tanzania (United Republic of)"] <- "Tanzania"
un_edu_data$location[un_edu_data$location == "Syrian Arab Republic"] <- "Syria"
un_edu_data$location[un_edu_data$location == "Syrian Arab Republic"] <- "Syria"
un_edu_data$location[un_edu_data$location == "Korea (Republic of)"] <- "South Korea"
un_edu_data$location[un_edu_data$location == "Russian Federation"] <- "Russia"
un_edu_data$location[un_edu_data$location == "Palestine, State of"] <- "Palestine"
un_edu_data$location[un_edu_data$location == "Moldova (Republic of)"] <- "Moldova"
un_edu_data$location[un_edu_data$location == "Lao People's Democratic Republic"] <- "Laos"
un_edu_data$location[un_edu_data$location == "Iran (Islamic Republic of)"] <- "Iran"
un_edu_data$location[un_edu_data$location == "Hong Kong, China (SAR)"] <- "Hong Kong"
un_edu_data$location[un_edu_data$location == "Eswatini (Kingdom of)"] <- "Eswatini"
un_edu_data$location[un_edu_data$location == "Congo (Democratic Republic of the)"] <- "Democratic Republic of Congo"
un_edu_data$location[un_edu_data$location == "Côte d'Ivoire"] <- "Cote d'Ivoire"
un_edu_data$location[un_edu_data$location == "Cabo Verde"] <- "Cape Verde"
un_edu_data$location[un_edu_data$location == "Brunei Darussalam"] <- "Brunei"
un_edu_data$location[un_edu_data$location == "Bolivia (Plurinational State of)"] <- "Bolivia"
```
<div style="text-align: justify"> 
Approach was taken to join the dataset by considering the ISO code.
However, there is no ISO code given in the UN data set to map with OWID
data set. Hence, I consider country names, but there were country
names mismatch in the UN and OWID dataset. In order, to rectify it I
replace the country names of the UN dataset to match with the OWID
dataset so I can join it to extract the education index for our
analysis. There are 18 entries with country names mismatch. Hence, I
map them to correct country name.
</div>
```{r}
#Joining UN and OWID datasets for extracting education index:
data_owid <- left_join(data_owid, un_edu_data, by = c('location'))
head(data_owid,2)
```

Education index is added as a new column to the OWID dataset.

```{r}
#Final check of missing values:
apply(is.na(data_owid),2, sum)
```

There is no any NA values present in our dataset. Our code is clean now
for further analysis.

## Exploratory Data Analysis

### Statistical summary

```{r}
#Summary statistics with SKIM function
skim(data_owid)
```
<div style="text-align: justify"> 
**SKIM function** provides an overview of the tidy dataset I have for
analysis. After pre-processing the OWID dataset is reduced to 180
observations and 19 variables. This overview represents clearly the
column type and its frequency in our dataset. There are 3 column types
namely character, date and numeric. The Variable type and their
characters are also described here with the data distribution in a
series of histogram for each variable separately.
</div>
### Visualization

```{r fig.align="center",fig.width = 9,fig.height=6}
#World map of country with their Total death:
l <- list(color = toRGB("black"), width = 0.5) # light black boundaries
# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)
fig <- plot_geo(data_owid)
fig <- fig %>% add_trace(
  z = ~total_deaths, color = ~total_deaths, colors = 'Reds',
  text = ~location, locations = ~iso_code, marker = list(line = l)
)
fig <- fig %>% colorbar(title = 'Covid Death - count', tickprefix = '')
fig <- fig %>% layout(
  title = 'Covid Death - Worldwide',
  geo = g
)
fig
```

The world map portrays the total death cases in each of the country.

**Note:** I have taken the variable "total death" only for this
visualization, in order to show the intensity of the covid death across
countries.

```{r fig.align="center",fig.width = 9,fig.height=6}
#World map of country with thier HDI:
l <- list(color = toRGB("black"), width = 0.5) # light black boundaries
# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'robinson')
)
fig <- plot_geo(data_owid)
fig <- fig %>% add_trace(
  z = ~human_development_index, color = ~human_development_index, colors = 'Greens',
  text = ~location, locations = ~iso_code, marker = list(line = l)
)
fig <- fig %>% colorbar(title = 'Human Development Index', tickprefix = '')
fig <- fig %>% layout(
  title = 'Human Development Index - Worldwide',
  geo = g
)
fig
```

The world map illustrates the HDI of each country.

```{r fig.align="center",fig.width = 9,fig.height=6}
#Scatterplot Matrices:
scatter_owid <- data_owid[c(8,14,16,15,19)] #Selecting only the required variables
scatter_owid.r <- abs(cor(scatter_owid)) #Getting the correlation
scatter_owid.col <- dmat.color(scatter_owid.r) #Getting colours for variables
scatter_owid.o <- order.single(scatter_owid.r) 
cpairs(scatter_owid, scatter_owid.o, panel.colors=scatter_owid.col, gap=.5,
       main="Scatterplot - Total death per million, HDI and its components" )
```
<div style="text-align: justify"> 
The required variables which corresponds to HDI and total death per
million is selected to plot the scatter plot. This illustrates how the
data is plotted against the other variable. I can clearly see that when
there is increase in HDI, life expectancy, and education index there is
increase in total death as well.
</div>
```{r fig.align="center",fig.width = 9,fig.height=6,warning=FALSE}
#scatterplot3d
attach(data_owid) 
scatterplot3d(median_age,human_development_index,total_deaths_per_million, 
              pch=16, 
              highlight.3d=TRUE,
              type="h", 
              main="3D Scatterplot",
              phi = 0,
              ticktype = "detailed"
              )
```
<div style="text-align: justify"> 
A 3D scatter plot is plotted with numerical variables. Here x and y axis
is taken as median_age and human_development_index with respect to our
target variable in the z axis which is total_deaths_per_million. Here I
can see all the data points are rising with increase in total deaths
which also denotes a positive correlation between them. Here we can
observe that as median age increases the HDI increases. This also
indicates that total death cases increases as the HDI increases.
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
#Checking the data distribution - quantile-quantile plot
DataExplorer::plot_qq(data_owid[4:10])
DataExplorer::plot_qq(data_owid[11:16])
```
<div style="text-align: justify"> 
Here I check the distribution of our data with the help of QQ plot. If
the data is normally distributed, the points in the QQ-normal plot lie
on a straight diagonal line. According to the qq plot the following
variables are normally distributed: Aged_65_older Aged_70_older
Cardiovasc_deaths_rates Diabetes_prevalence Life expectancy
Human_Development_Index Total_deaths_per_million
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
#Scatterplot - HDI with total death with respect to each Continent
ggplot(data = data_owid) + 
  geom_point(mapping = aes(x = human_development_index, y = total_deaths_per_million, color = continent))
```
<div style="text-align: justify"> 
This scatter plot is plotted to see the relationship of
total_deaths_per_million and human_development_index with respect to
each continent. Here I can observe that Europe has high HDI and it has
contributed to more death. In contrast, I have Africa which has low HDI
and low death. Continent like Asia, South America, North America and
Oceania are moderate in HDI and total death.
</div>
```{r fig.align="center",fig.width = 16,fig.height=8}
#Top 5 countries with high HDI and low 5 countries with less HDI
#select relevant variables for plotting
Owid_data_Graph <- data_owid %>% 
  select(iso_code,location,median_age,aged_70_older,cardiovasc_death_rate,diabetes_prevalence,life_expectancy,gdp_per_capita,education_index,human_development_index,total_deaths_per_million)
OWiD_hdi <- Owid_data_Graph %>%
  select(location,human_development_index)
#Select top5 and bottom 5 countries by HDI
OWiD_high_hdi <- OWiD_hdi  %>% top_n(5)
OWiD_low_hdi <- OWiD_hdi  %>% top_n(-5)
#Combine top and bottom HDI countries
OWiD_high_low<-bind_rows(OWiD_high_hdi, OWiD_low_hdi)
#Color for bar chart
coul <- brewer.pal(5, "Set2") 
barplot(OWiD_high_low$human_development_index,
        main = "Top 5 and low 5 HDI Countries",
        ylab = "HDI",
        xlab = "Countries",
        ylim=c(0,1),
        names.arg = OWiD_high_low$location,
        col = coul,
        width=c(1),
        density=c(20),
        las=1,
        horiz = FALSE
)
```
<div style="text-align: justify"> 
The above bar plot illustrates the countries which has high HDI and low
HDI. Hong Kong, Iceland, Ireland, Norway and Switzerland are countries
with high HDI. Burundi, Central African Republic, Chad, Niger and South
Sudan are countries with low HDI.
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
#Top 5 countries (HDI components with total death per million)
#High HDI countries, plot all HDI parameters
#Select - Top5 HDI countries
OWiD_all_Param_Top <- Owid_data_Graph[order(-Owid_data_Graph$human_development_index),][1:5,]
#apply log to scale down the index values
OWiD_all_Param_Top$total_deaths_per_million <- round(log(OWiD_all_Param_Top$total_deaths_per_million),2)
OWiD_all_Param_Top$gdp_per_capita <- round(log(OWiD_all_Param_Top$gdp_per_capita),2)
OWiD_all_Param_Top$life_expectancy <- round(log(OWiD_all_Param_Top$life_expectancy),2)
OWiD_all_Param_Top$education_index <- round(log(OWiD_all_Param_Top$education_index),2)
OWiD_all_Param <- OWiD_all_Param_Top  %>%
  select(location,life_expectancy,gdp_per_capita,education_index,total_deaths_per_million)
#convert to Pivot longer for plot
OWiD_all_Param_PL <- OWiD_all_Param %>%
  pivot_longer(cols=-c("location"), names_to = "Index" , values_to = "Log_Index")
ggplot_all_param <- ggplot(data=OWiD_all_Param_PL, aes(x=location, y=Log_Index, fill=Index)) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label=Log_Index), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  scale_fill_manual(values = coul)+
  theme_minimal()
ggplot_all_param + ggtitle("Top 5 HDI countries for HDI Parameters")
```
<div style="text-align: justify"> 
I plot a bar graph taking country with high HDI to visualize how
life_expectancy, gdp_per_capita and education_index care are with
respect to the total death. Here we can observe that all the components
plotted here are relatively similar, except for Ireland and Switzerland
which shows slight increase in deaths despite being comparatively
similar to other countries.

**Note**: Applied log for all variables to fit them into scale for the
bar chart visualization.
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
#Low 5 countries (HDI components with total death per million)
#Select - Low 5 HDI countries
OWiD_all_Param_Top <- Owid_data_Graph[order(Owid_data_Graph$human_development_index),][1:5,]
#apply log to scale down the index values
OWiD_all_Param_Top$total_deaths_per_million <- round(log(OWiD_all_Param_Top$total_deaths_per_million),2)
OWiD_all_Param_Top$gdp_per_capita <- round(log(OWiD_all_Param_Top$gdp_per_capita),2)
OWiD_all_Param_Top$life_expectancy <- round(log(OWiD_all_Param_Top$life_expectancy),2)
OWiD_all_Param_Top$education_index <- round(log(OWiD_all_Param_Top$education_index),2)
OWiD_all_Param <- OWiD_all_Param_Top  %>%
  select(location,life_expectancy,gdp_per_capita,education_index,total_deaths_per_million)
#convert to Pivot longer for plot
OWiD_all_Param_PL <- OWiD_all_Param %>%
  pivot_longer(cols=-c("location"), names_to = "Index" , values_to = "Log_Index")
ggplot_all_param <- ggplot(data=OWiD_all_Param_PL, aes(x=location, y=Log_Index, fill=Index)) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label=Log_Index), vjust=1.6, color="white", 
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  scale_fill_manual(values = coul)+
  theme_minimal()
ggplot_all_param + ggtitle("Low 5 HDI countries for HDI Parameters")
```
<div style="text-align: justify"> 
I plot a bar graph taking country with high low to visualize how
life_expectancy, gdp_per_capita and education_index care are with
respect to the total death. Here we can observe that Central African
Republic has contributed to more death even though their other
components are relatively similar to other countries and Burundi has
less deaths.

**Note**: Applied log for all variables to fit them into scale for the
bar chart visualization.
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
# Top 5 and low 5 countries with high cardiac death rate and their covid death
#Select - Top 5 HDI countries
OWiD_all_cardiac_top <- Owid_data_Graph[order(-Owid_data_Graph$human_development_index),][1:5,]
#select required variables cardiac_death and covid death
OWiD_all_cardiac_covid_high <- OWiD_all_cardiac_top  %>%
  select(location,cardiovasc_death_rate,total_deaths_per_million,human_development_index)
#Select - bottom 5 HDI countries
OWiD_all_cardiac_bottom <- Owid_data_Graph[order(Owid_data_Graph$human_development_index),][1:5,]
#select required variables cardiac_death and covid death
OWiD_all_cardiac_covid_bottom <- OWiD_all_cardiac_bottom  %>%
  select(location,cardiovasc_death_rate,total_deaths_per_million,human_development_index)
#Combine top and bottom HDI countries
OWiD_all_cardiac_top<-bind_rows(OWiD_all_cardiac_covid_high, OWiD_all_cardiac_covid_bottom)
#Apply log to to fit to the bar chart scale
OWiD_all_cardiac_top$cardiovasc_death_rate <- round(log(OWiD_all_cardiac_top$cardiovasc_death_rate),2)
OWiD_all_cardiac_top$total_deaths_per_million <- round(log(OWiD_all_cardiac_top$total_deaths_per_million),2)
#Convert to Pivot longer for plot
OWiD_all_cardiac_top <- OWiD_all_cardiac_top %>%
  pivot_longer(cols=-c("location"), names_to = "Index" , values_to = "value")
#Horizontal Plot
ggplot(OWiD_all_cardiac_top[order(OWiD_all_cardiac_top$Index, decreasing = T),],
       aes(value,  y=reorder(location, -value), label= round(value,2), fill=factor(Index, levels=c("total_deaths_per_million","cardiovasc_death_rate","human_development_index")))) + 
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Category", labels = c("total_deaths_per_million","cardiovasc_death_rate","human_development_index"))+
  scale_x_continuous(breaks=c(2,4,6,8,10,12,14)) +
  ylab("Countries") + 
  ggtitle("Top 5 and bottom 5 HDI countries with corresponding cardiac death rate and covid_death")
```
<div style="text-align: justify"> 
Seeing the trend of the top 5 and bottom 5 countries on HDI and their
cardiac death and covid death, there is a visible trend which shows,

a) for the high HDI countries, the cardiac death is relatively low but
the covid death is relatively high.

b) for the low HDI countries, the cardiac death is relatively high but
the covid death is relatively low.

**Note**: Applied log to cardiac death rate and covid death variable to
fit them in the bar chart. This chart is to show the the 3 variables and
their trend, not the actual value.
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
# Top 5 and low 5 countries with high diabetes prevalence and their covid death
#Select -  Top 5 HDI countries
OWiD_all_diabetes_top <- Owid_data_Graph[order(-Owid_data_Graph$human_development_index),][1:5,]
#select required variables cardiac_death and diabetes prevalence
OWiD_all_diabetes_top <- OWiD_all_diabetes_top  %>%
  select(location,diabetes_prevalence,total_deaths_per_million,human_development_index)
#Select - bottom 5 HDI countries
OWiD_all_diabetes_bottom <- Owid_data_Graph[order(Owid_data_Graph$human_development_index),][1:5,]
#select required variables cardiac_death and diabetes prevalence
OWiD_all_diabetes_bottom <- OWiD_all_diabetes_bottom  %>%
  select(location,diabetes_prevalence,total_deaths_per_million,human_development_index)
#Combine top and bottom HDI countries
OWiD_all_diabetes<-bind_rows(OWiD_all_diabetes_top, OWiD_all_diabetes_bottom)
#Apply log to to fit to the bar chart scale
OWiD_all_diabetes$total_deaths_per_million <- round(log(OWiD_all_diabetes$total_deaths_per_million),2)
#Convert to Pivot longer for plot
OWiD_all_diabetes <- OWiD_all_diabetes %>%
  pivot_longer(cols=-c("location"), names_to = "Index" , values_to = "value")
#Horizontal Plot
ggplot(OWiD_all_diabetes[order(OWiD_all_diabetes$Index, decreasing = T),],
       aes(value,  y=reorder(location, -value), label= round(value,2), fill=factor(Index, levels=c("diabetes_prevalence","total_deaths_per_million","human_development_index")))) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Category", labels = c("diabetes_prevalence","total_deaths_per_million","human_development_index"))+
  scale_x_continuous(breaks=c(2,4,6,8,10,12,14,16)) +
  ylab("Countries") + 
  ggtitle("Top 5 and bottom 5 HDI countries with corresponding diabetes prevalance and covid_death")
```
<div style="text-align: justify"> 
Diabetes prevalence for high and low HDI countries are not hugely
varied. No clear trend been seen between diabetic prevalence and covid
death for high and low HDI countries. Hence the covid death may not be
related to diabetes prevalence, but need to be analysed with other
related parameters.

**Note**: Applied log to covid death variable to fit all variables in
the bar chart. This chart is to show the the 3 variables and their
trend, not the actual value.
</div>
```{r fig.align="center",fig.width = 12,fig.height=6}
# Top 5 and bottom 5 countries with Aged-70_older and their median age and death per million
#Select - 5 aged_70_older
OWiD_all_aged_top <- Owid_data_Graph[order(-Owid_data_Graph$human_development_index),][1:5,]
#select required variables cardiac_death and diabetes prevalence
OWiD_all_aged_top <- OWiD_all_aged_top  %>%
  select(location,median_age,aged_70_older,total_deaths_per_million,human_development_index)
#Select - bottom 5 aged_70_older
OWiD_all_aged_bottom <- Owid_data_Graph[order(Owid_data_Graph$human_development_index),][1:5,]
#select required variables cardiac_death and diabetes prevalence
OWiD_all_aged_bottom <- OWiD_all_aged_bottom  %>%
  select(location,median_age,aged_70_older,total_deaths_per_million,human_development_index)
#Combine top and bottom aged_70_older
OWiD_all_aged<-bind_rows(OWiD_all_aged_top, OWiD_all_aged_bottom)
#Apply log to to fit to the bar chart scale
OWiD_all_aged$total_deaths_per_million <- round(log(OWiD_all_aged$total_deaths_per_million),2)
#Convert to Pivot longer for plot
OWiD_all_aged <- OWiD_all_aged %>%
  pivot_longer(cols=-c("location"), names_to = "Index" , values_to = "value")
#Horizontal Plot
ggplot(OWiD_all_aged[order(OWiD_all_aged$Index, decreasing = T),],
       aes(value,  y=reorder(location, -value), label= round(value,2), fill=factor(Index, levels=c("aged_70_older","median_age","total_deaths_per_million","human_development_index")))) + 
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Category", labels = c("aged_70_older","median_age","total_deaths_per_million","human_development_index"))+
  scale_x_continuous(breaks=c(10,20,30,40,50,60,70,80)) +
  ylab("Countries") + 
  ggtitle("Top 5 and bottom 5 HDI countries with corresponding aged_70_order and median age and total deaths per million")
```
<div style="text-align: justify"> 
Median age and aged_70_older are related variables is a known fact. From
this bar chart we can infer that, higher the HDI means Higher the median
age and higher the covid death.

**Note:** Applied log to covid death variable to fit all variables in
the bar chart scale. This chart is to show the the 4 variables and their
trend, not the actual value.
</div>

### Correlation

```{r}
owid_corr <- cor(data_owid[,c(4:16)])
corrplot(owid_corr, method = "circle")
```
<div style="text-align: justify"> 
Correlation defines how strong the variables are related to each other.
There are 3 types of correlation,

1\. Positive correlation

2\. Negative correlation

3\. Zero correlation

Correlation coefficient is a statistical measure to identify the type of
correlation existing within the variables. From the correlation plot. We
can observe from the correlation plot that median_age, aged_65_older and
aged_70_older are more correlate to each other and also shows a high
positive correlation. Female_smokers also shows a high positive
correlation with variables related to age. cardiovasc_death_rate shows
negative correlation. However, they are low so it can be ignored.
</div>
# Analysis

## Hypothesis testing
<div style="text-align: justify"> 
Hypothesis testing is a statistical test performed to prove an idea or
an assumption by using sample data. The test provides evidence
concerning the plausibility of the hypothesis. The motive of the project
is to identify the correlation between Human Development Index and the
total deaths caused by COVID-19 pandemic. This can be found by
performing the hypothesis testing. The selected variables are
quantitative data, hence we can perform a **Spearman's correlation**
test which provides us correlation as well as the direction of the
relationship between two continuous variables.

[**Hypothesis Statement:**]{.ul}

**H0 (Null hypothesis) :** There is no linear relationship between HDI
and number of death

**H1 (Alternate hypothesis) :** There is a linear relationship between
HDI and number of death

The H0 and H1 are mutually exclusive, and only one can be true. However,
one of the two hypothesis will always be true

It is required to check the normality of the data before a hypothesis is
performed.Here I have chosen Shapiro-Wilk normality test for evaluation
as they are widely recommended and more powerful than the other tests.
Shapiro-Wilk normality testdepends on the correlation between the data
and their normal scores.
</div>

**Shapiro-Wilk normality test**

```{r}
#Normality test - Shapiro-Wilk
shapiro.test(data_owid$human_development_index)
#p-value =  1.416e-05
shapiro.test(data_owid$total_deaths_per_million)
#p-value = 5.283e-12
```
<div style="text-align: justify"> 
From the output, the two p-values are less than the significance level
of 0.05 implying that the distribution of the data is significantly
different from normal distribution, which means the data is not normally
distributed.
</div>

**Hypothesis testing:**

1.  **HDI vs Total Death - All countries**
<div style="text-align: justify"> 
I perform Spearman's correlation test because from the Shapiro-Wilk
normality test it can be observed that the data does not follow normal
distribution.
</div>
```{r}
hdi_corr_all <- cor.test(data_owid$human_development_index, data_owid$total_deaths_per_million, 
                     method = "spearman", exact = FALSE)
hdi_corr_all
```
<div style="text-align: justify"> 
The p-value here is 2.2e-16, which is less than the significance level
alpha = 0.05. From this it can concluded that HDI and total deaths are
significantly correlated with a correlation coefficient of 0.6109514.
Hereby, we reject H0 and accept H1.
</div>
2.  **HDI vs Total Death - High HDI and low HDI:**
<div style="text-align: justify"> 
Assumption : Countries with HDI 0.7 and above are considered to be
countries with high HDI Countries with HDI 0.5 and below are considered
to be countries with low HDI
</div>
```{r}
#Hypothesis Test - Country with high HDI vs Total Death
#Selecting countries with high HDI
OWiD_high_hdi <- data_owid %>% filter(human_development_index > 0.7)
OWiD_high_hdi
#Spearman's correlation test:
hdi_corr_high <- cor.test(OWiD_high_hdi$human_development_index, OWiD_high_hdi$total_deaths_per_million, 
                     method = "spearman", exact = FALSE)
hdi_corr_high
```
<div style="text-align: justify"> 
Here p-value = 0.5004 and Correlation coefficient = 0.6403101. Here we
fail to reject H0. It can be concluded that there is no linear relation
between HDI and number of deaths in countries with high HDI.
</div>
3.  **HDI vs Total Death - Country with low HDI vs Total Death**

```{r}
#Selecting countries with low HDI
OWiD_low_hdi <- data_owid %>% filter(human_development_index < 0.5)
OWiD_low_hdi
#Spearman's correlation test :
hdi_corr_low <- cor.test(OWiD_low_hdi$human_development_index, OWiD_low_hdi$total_deaths_per_million, 
                     method = "spearman", exact = FALSE)
hdi_corr_low
```

Here p-value = 0.0003852 and Correlation coefficient = 0.7453452 p value
is less than 5%, hence we reject H0 and accepting H1 which states there
is linear relation between HDI and total death in countries with low HDI

4.  **HDI vs Total Death - Continent Wave 1 - European region**

```{r}
#Extracting data with date of Wave 1
data_owid_wave1 <- data_Covid %>% 
  select(continent, location, date, human_development_index, total_deaths_per_million) %>% 
  group_by(location) %>% filter(date == "2020-06-01") %>% # Take the most recent date per country
  group_by() #removing groupby
```

Creating another dataset filtering date of 2020-06-01 which denotes the
first wave of the pandemic.

```{r}
#Renaming the variables:
colnames(data_owid_wave1)[6] <- "total_deaths_per_million_Wave1"
head(data_owid_wave1,1)
```

Rename the column - total_deaths_per_million_Wave as
total_deaths_per_million_Wave1 which would be used in further analysis
while joining both the dataset - data_owid and data_owid_wave1

```{r}
#Joining both the dataset
data_owid_wave1 <- left_join(data_owid, data_owid_wave1, by = c('location'))
head(data_owid_wave1,2) 

apply(is.na(data_owid_wave1),2, sum)
```

19 countries does not have data as of 2020-06-01. This might be a reason
that wave 1 would not have started for these countries as of June 2020.

```{r}
#Removing NA from total_deaths_per_million.y
data_owid_wave1 <- data_owid_wave1 %>% filter(!is.na(total_deaths_per_million.y))
```

```{r}
#Filtering continent - Europe
data_owid_wave1 <- data_owid_wave1 %>% 
  filter(continent.x == "Europe")
#Spearman's correlation test:
hdi_corr_europe <- cor.test(data_owid_wave1$human_development_index.y, data_owid_wave1$total_deaths_per_million.y,
                            method = "spearman", exact = FALSE)
hdi_corr_europe
```
<div style="text-align: justify"> 
Here p-value = 0.02015 and Correlation coefficient = 0.463047,
here p value is less than 5%, hence we reject H0 and accepting H1 which
states there is linear relation between HDI and total death in European
countries when first wave occurred.
</div>

### Hypothesis test - Table and comparison

```{r}
#Creating Matrix to add correlation result:
table_corr<- matrix(0,4,4) # entry, rows, columns
row.names(table_corr) <- c("All_countries", "Countries_High_HDI ", "Countries_low_HDI ", "Wave_1_Europe")
colnames(table_corr) <- c("P-value","T-test","Correlation", "H0/H1")

#Result for all countries
table_corr[1,1] <- round(hdi_corr_all$p.value,3)
table_corr[1,2] <- round(hdi_corr_all$statistic,3)
table_corr[1,3] <- round(hdi_corr_all$estimate,3)
table_corr[1,4] <- "H1"

#Result for High HDI countries
table_corr[2,1] <- round(hdi_corr_high$p.value,3)
table_corr[2,2] <- round(hdi_corr_high$statistic,3)
table_corr[2,3] <- round(hdi_corr_high$estimate,3)
table_corr[2,4] <- "H0"

#Result for Low HDI countries
table_corr[3,1] <- round(hdi_corr_low$p.value,3)
table_corr[3,2] <- round(hdi_corr_low$statistic,3)
table_corr[3,3] <- round(hdi_corr_low$estimate,3)
table_corr[3,4] <- "H1"

#Result for wave1 - Europe
table_corr[4,1] <- round(hdi_corr_europe$p.value,3)
table_corr[4,2] <- round(hdi_corr_europe$statistic,3)
table_corr[4,3] <- round(hdi_corr_europe$estimate,3)
table_corr[4,4] <- "H1"

table_corr <- as.data.frame(table_corr)
```

### Summarizing the hypothesis results:

```{r}
table_corr
```

Plotting the correlation obtained in the hypothesis results:

```{r}
#human_development_index and total_deaths_per_million - All countries
ggscatter(data_owid, x = "human_development_index", y = "total_deaths_per_million", 
          add = "reg.line", conf.int = TRUE, 
          color = "blue",
          palette = NULL,
          cor.coef = TRUE, 
          cor.method = "spearman",
          xlab = "Human Development Index", ylab = "Deaths per million",
          title = 'Human Development Index and Deaths per million - All countries')
```
<div style="text-align: justify"> 
We can observe from the Spearman's Correlation test between the HDI and
total deaths in all country that H0 is rejected since p-value is less
than significance value of 0.05. This brings to a conclusion that both
the variables are linearly dependent. Furthermore, by visualizing the
result through scatter plot, we can conclude saying the variables are
positively correlated. The Spearman's correlation coefficient r which is
0.61 which again justifies that it is far away from 0 which makes the
linear relationship more stronger between the HDI and total deaths. It
can be interfered that as HDI increases the total death also increases.
</div>

```{r}
#Plot for High HDI:
ggscatter(OWiD_high_hdi, x = "human_development_index", y = "total_deaths_per_million", 
          add = "reg.line", conf.int = TRUE, 
          color = "red",
          palette = NULL,
          cor.coef = TRUE, 
          cor.method = "spearman",
          xlab = "Human Development Index", ylab = "Deaths per million",
          title = 'Human Development Index and Deaths per million - High HDI countries')
```
<div style="text-align: justify"> 
The Spearman's Correlation test between the country with high HDI and
total deaths concludes that H1 is accepted and we failed to reject H0
since p-value is more than significance value of 0.05. This concludes
that both the variables are not linearly dependant. With the visualizing
of scatter plot, we can conclude saying the variables are not correlated
with each other and they do not project any linear dependencies between
them. To makes this statement even stronger, the Spearman's correlation
coefficient r which is 0.064 which displays zero correlation between the
two variables.
</div>
## Linear Regression - Modelling:
<div style="text-align: justify"> 
From the hypothesis testing, we can see there is no relation between the
total deaths per million and the HDI in countries with high HDI. Hence
to identify what are the other contributing factors leading to covid
death, I now perform a linear regression between countries with high
HDI and covid deaths.
</div>
```{r}
#Feature selection:
#Selecting only required data as per correlation plot:
data_owid <- data_owid %>%
            filter(human_development_index > 0.7) %>%
            select (population,  median_age, gdp_per_capita, cardiovasc_death_rate, 
                    diabetes_prevalence, male_smokers,hospital_beds_per_thousand, life_expectancy, 
                    human_development_index, education_index, total_deaths_per_million)
            
```

Select only required variables to see the relationship between
death and HDI.

```{r}
#Normalise the data points:
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}
data_owid$population<-normalize(data_owid$population)
data_owid$median_age<-normalize(data_owid$median_age)
data_owid$gdp_per_capita<-normalize(data_owid$gdp_per_capita)
data_owid$cardiovasc_death_rate<-normalize(data_owid$cardiovasc_death_rate)
data_owid$diabetes_prevalence<-normalize(data_owid$diabetes_prevalence)
data_owid$male_smokers<-normalize(data_owid$male_smokers)
data_owid$hospital_beds_per_thousand<-normalize(data_owid$hospital_beds_per_thousand)
data_owid$life_expectancy<-normalize(data_owid$life_expectancy)
data_owid$human_development_index<-normalize(data_owid$human_development_index)
data_owid$education_index<-normalize(data_owid$education_index)
```

Since the dataset is not normalized as per the qq plot, performing a
normality function to normalize data, which is essential for performing
linear regression.

```{r}
#Splitting the data
set.seed(1234)
create_train_test <- function(data, size = 0.8, train = TRUE) {
  n_row = nrow(data)
  total_row = size * n_row
  train_sample <- 1: total_row
  if (train == TRUE) {
    return (data[train_sample, ])
    } else {
      return (data[-train_sample, ])
    }
  }
data_train <- create_train_test(data_owid, 0.8, train = TRUE)
data_test <- create_train_test(data_owid, 0.8, train = FALSE)
dim(data_train)
dim(data_test)
```

Here the dataset is split into two datasets. 1. Train 2. Test

```{r}
#Fit, train and predicting the model
fitControl <- trainControl(method = "none")
lm_train_pred <- train(total_deaths_per_million ~., 
                       data = data_train,
                       method = "lm",
                       trControl = fitControl)
lm_train_pred
lm_pred = predict(lm_train_pred, data_test)
summary(lm_train_pred)
```
<div style="text-align: justify"> 
From the linear regression summary it can be confirmed
that

a\. Cardiac_death_rate and male_Smokers are not statistically
significant because their p-values are greater than the usual
significance level of 0.05

b\. The HDI coefficient in the regression equation is 2137.95. This
coefficient represents the mean increase of total deaths per million for
every additional one increase in HDI, which means if HDI increases by
0.1, the average total deaths per million increases by 2137.95. It is
also interesting to find that median_age shows a high positive
correlation with covid death. The coefficient of median age in the
regression equation is 2572.14. If median age increases by 1, the average
total deaths per million increases by 2572.14.

c\. There is a negative coefficient for variables such as
gdp_per_capita, diabetes_prevalence, life_expectancy and education_index
which suggests that as these variables increases, the covid death
decreases.
</div>
# Conclusion
<div style="text-align: justify"> 
From the overall analysis, identified that there exist a linear
relation between HDI and covid death. However, while exploring deep into
separate countries with high HDI and low HDI we can observe that, high HDI
countries has no relationship with covid death. Hence performing a
linear relation to identify what are the other factors contributing to
death in countries with high HDI. It can be concluded from the linear
model that median age variable (aged population) is one of the prominent
factor which contributes to total death, which means high median age
contributes to high covid deaths.
</div>
# Critique
<div style="text-align: justify"> 
**Alternative Explanations:**

From general observation, another probable reason why countries with
high HDI had high death rates is due to the longer life expectancy as
compared to low HDI countries. This means High HDI countries have a
large population of elederly people which are more prone to risk of
contracting diseases due to weakened immune systems thus the corona
virus pandemic took a toll on high HDI countries. We can also notice
that when vaccines where being administered, high priority was given to
the old age since they were more at risk.

**Limitations:**

1.Initially OWID dataset has 237 observations with 20 variables.
However, after cleaning the data with several pre-processing steps and
handling all missing values the dataset was reduced to 180 observations
and 19 variables which is a small dataset. Around 24% of the data is
been removed, which might have an impact on the analysis and statistical
output.

2.No scientific methods applied other than mean and median to replace NA
values.
</div>